A multi-scale story of the diffusion of a new technology: the web

### 2025-03-12 paper/submission2/response.qmd

> This paper offers such a contribution: an economic geography study
revealing the importance of spatial mechanisms for the diffusion of a
digital technology that is the Web in the UK at a high level
of spatial granularity 
from the very early days until its maturity (1996-2012).
It specifically focuses on the diffusion of commercial websites.
It does so by employing a novel source of big data and machine learning
algorithms to model the *active* engagement with Web -- that is creating and 
maintaining instead of just browsing a website -- during that period
and tests how different spatial and temporal mechanisms shaped its
diffusion. This paper also exemplifies how the combination of data
sources and state-of-the-art research methods which escape the
traditional social science domain can offer new lenses to geographical
research regarding the understanding of technological diffusion.

@@

\color{teal}
\noindent 4. Motivation: The motivation for the study is unclear. The text raises several issues, but it is unclear which are specific to digital technologies and which pertain to technological diffusion more broadly.
\color{black}

There are three paragraphs in the revised introduction that highlight the 
paper's motivation. The first paragraph highlights the importance of this work 
from a stakeholder perspective. The second emphasises the importance of 
technology and its adoption in economic geography and macroeconomic growth 
theories. The third one, motivates the paper from the standpoint of digital 
geographies and connects it to that literature.

> <!-- motivation: stakeholders -->
The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffuse over space and time and use this knowledge to make predictions
about the diffusion of related future technologies. As per
@leibowicz2016representing, historical studies agree that technologies
diffuse differently in terms of times, rates, and geographies and can be
driven by related policies [@victor1993]. @meade2021modelling highlight
that a variety of actors have a direct interest in gaining such
knowledge including network equipment suppliers, network operators,
regulatory and local authorities. These processes and their effects vary
a lot across scales: although the diffusion of a new technology might
not be optimal at a local level, it might be beneficial from a global
perspective as it could lead to faster diffusion to less advantaged
places [@leibowicz2016representing]. Despite the spatial heterogeneity
of such diffusion mechanisms and the policy relevance, there is a
scarcity of studies analysing the diffusion of new technologies
at a detailed geographical level because of the lack of detailed enough
data and adequate research methods.

<!-- motivation: regional development -->
>This gap becomes even more prominent if we consider the importance of
technology and its adoption in economic geography and macroeconomic
growth theories. Simply put, differences in technological adoption and
sophistication may lead to uneven development as new technologies can
increase productivity and allow firms to introduce new products
[@solow1957technical; @aghion1990model; @kemeny2011international].
Linked to the subject of this paper, the Web and, more specifically, 
commercial websites can be considered as the underpinning
technology for the transformative process identified by
@capello2024nexus as the *digital service economy*. Importantly, the
focus here is not the Web in general, but instead commercial websites.
After all, diffusion together with invention and innovation are
considered the pillars of technological change [@das2022diffusion].
Findings from previous studies illustrated that early engagement with
digital technologies such as the Web led to longstanding positive
productivity gains for these regions that hosted early adopters
[@tranosuk]. Hence, obtaining a robust understanding of the spatial
diffusion mechanisms is crucial in mapping future development trajectories.

<!-- NEW motivation: intangible/digital -->
> Another line of motivation has to do with the very nature of the Web as a 
digital technology and how it evolved over time. The digital geographies field 
has debunked premature claims about the Internet's placeless-ness by mapping 
various of its facets and technological layers [@tranos2013death], but the lack 
of relevant long time-series data prevented researchers from observing digital
phenomena overtime. Most of these studies focused on its hard infrastructure 
[e.g. @malecki2002economic; @moss2000internet], on Internet subscribers 
[e.g. @blank2018local], or on social media users [e.g. @crampton2013beyond].
Exceptions include early work from @zook2000web, @zook2000web and 
@moss1997tracking, which utilised domain name registration data and use it to
map and illustrate the spatial footprint of the web content. Pivotal as these 
contributions have been in understanding the economic geographies of the 
Internet, they were based on cross-sectional, often spatially detailed, mapping 
and on short temporal comparisons (e.g. 1993-1998) and did not observe the 
evolution and diffusion of the Web until its maturity. As we know from the 
relevant literature, different technologies diffuse at different rates and 
there is no theory that can explain these differences 
[@leibowicz2016representing]. Hence, mapping and modelling the diffusion of the
Web can offer useful insights.

@@

\color{teal}
\noindent 7. Recommended Structure: To align with the common structure in Journal of Economic Geography (JoEG), consider splitting the introduction into a shorter introduction and a dedicated theory section merged with the literature review.
\color{black}

The introduction is now shorter and the literature review section has been 
re-structured and re-named as "The mechanisms that shape the diffusion of a new 
technology".

\color{teal}
\noindent 3. Literature Review / Theory\
\noindent 1. Synthesis and Motivation: The literature review should be transformed into a theory section. Rather than presenting a historical review, synthesize the arguments to clearly motivate and inform your hypotheses.
\color{black}

This section has been re-written and structured around the two hypotheses:

> H~1~: The three diffusion mechanisms namely distance, urban hierarchy
and the S-shaped pattern of the cumulative level of adoption shaped
the diffusion of the commercial Web in the UK even at granular geographical 
scales.

> H~2~: The diffusion of the commercial Web is shaped by both early adoption 
and latecomer effects.

@@

\color{teal}
\noindent 2. Knowledge Gaps: The current review does not clearly identify the knowledge gaps that this study aims to address. Clarify whether the goal is to confirm known processes with new data and methods or to explore unique diffusion patterns related to websites.
\color{black}

The knowledge gaps have been explicitly highlighted in the restructured 
literature review section. For instance:

> Although the above processes have been studied in innovation studies, the distancing of economic geographers from research questions about the diffusion of new technologies and innovations resulted to a knowledge gap regarding the explanatory power of these mechanisms at granular, local scales [@ding2010modeling].

> Although the early adoption and leapfrogging literature has not paid much attention on cities and regions [@yu2018sustainability], the potential local economic benefits justify the need to investigate the dynamics of the web diffusion in different scales
in the UK. Hence, the second empirical hypothesis states:

@@

\color{teal}
\noindent 3. Lack of Hypotheses: To clarify what the paper delivers and what knowledge gaps it fills, I strongly recommend developing a number of hypotheses that will be tested.
\color{black}

The following two hypotheses have been added and the whole paper is now 
structured around them:

> H~1~: The three diffusion mechanisms namely distance, urban hierarchy
and the S-shaped pattern of the cumulative level of adoption shaped
the diffusion of the commercial Web in the UK even at granular geographical 
scales.

> H~2~: The diffusion of the commercial Web is shaped by both early adoption 
and latecomer effects.

@@

\color{teal}
\noindent 4. Fit to title: The term “multi-scalar” appears in the title but is never really discussed in the paper.
\color{black}

This was discussed from the data section onwards, but indeed not before that. 
Section 2 includes a new paragraph describing the second hypothesis (H~2~), 
which explicitly discusses the value of the multiscale approach: 
"Understanding the diffusion of a new technology in multiple, detailed spatial
scales is of interest to economic geography as ..."

The two scales are firstly introduced in the introduction: "Importantly, the analysis adopts two distinct geographical scales: the LAD and the most detailed scale of UK census areas, the Output Areas (OA)".

@@

The order of presentation has been amended accordingly and the two scales of 
analyses are presented after the data.

\color{teal}
\noindent 2. Postcode Identification: The approach of identifying postcodes on websites is innovative, but potential errors need to be discussed. Specifically:
- How do you ensure that a postcode on a website accurately represents the 
company’s location or its subsidiary?
- How do you handle numbers that resemble postcodes but are not actual postcodes?
- What about references to events or other organizations unrelated to the website’s owner?
Include a brief discussion of these potential issues.
\color{black}

A number of different strategies have been developed to address the potential 
issues described above.

\noindent 1. To prevent false positives, the resulted postcodes
were matched against official postcode lookup files [@ons]. This was not 
explicitly mentioned in the first submission of the paper and the previous 
sentence has now been added in the relevant paragraph:

@@

> The text from these webpages was scanned using a regular expression
(regex) to identify strings of text which resemble UK postcodes and one
fifth of them included a mention to a postcode [@BL2013geo]. This
information allows the geolocation of the data and the creation of the
LAD and OA counts. To prevent false positives, the resulted postcodes
were matched against official postcode lookup files [@ons].

\noindent 2. Then, to remove the bias from link farms associated with a handful 
of postcodes having irregular high numbers of websites pointing to them, a 
targeted data imputation procedure was developed. This is detailed in the data 
section -- see also Figure 1.

\noindent 3. Regarding the points about subsidiary company location and event
postings, these cannot be directly observed and distinguished. However, a 
process that has been successfully used in previous studies has been adopted 
here. Firstly, given that the focus of the paper is commercial websites, the 
relevant literature illustrates that such websites serve specific purposes:

> ... they are used to diffuse information, support online transactions and share 
opinions [@THELWALL2000441; @blazquez2018big].

Secondly, the webpage-to-website aggregation process leads to websites with 
different number of postcodes as presented in Table 1. To decrease heterogeneity, 
the analysis firstly focuses on websites with one unique postcode. Given the the
commercial nature of these websites (.co.uk), it is fair to expect that such 
websites with one unique postcode in their web text represent companies with a 
single trading location. The analysis included in the main body of the paper is
based on this subset of the website population. For year 2000, this includes 
70% of all the archived, geolocated commercial websites. Then, as a robustness
check all the different steps of the analysis are repeated for websites with up 
to 10 postcodes and the results are presented in the appendix. This includes 94%
of all the archived, geolocated commercial websites for 2000. For this step, 
websites with multiple postcodes were counted multiple times, one for each 
different postcode. All the results are qualitatively the same and only 
marginally different, which is an indication of the robustness of the website 
geolocation strategy.

@@

As a further argument for the robustness of the above process, the results of a 
previous study which used the same approach for a small subset of these data 
from a specific neighbourhood in London matched in terms of accuracy previous 
research that was based on extensive qualitative work, interviews and 
observational studies [@shoreditch]. This last point has now been added in the 
relevant section:

> Previous research which used the same methodology for a small subset of these 
data illustrates the robustness of the above approach: its results matched in 
terms of accuracy past rounds of research based on extensive qualitative work, 
interviews and observational studies for the same study area [@shoreditch].

\color{teal}
\noindent 3. Equation 2: The rationale behind Equation 2 is unclear. Shouldn’t the primary predictor of website density at time tbe the location density at t-1? Why was this approach chosen over a growth model? Additionally, clarify how spatial dependencies are accounted for in the model.
\color{black}

Because of the restructuring of the methods section, which now includes all the 
details of the modelling exercise, Equation 2 is followed by Equation 3 that 
clarifies the spatial and temporal dependencies. All these details are described
in the following paragraph, which is placed just before Equation 3:

> To capture the hierarchical mechanism the model include as predictors a
one year lag of website density in London, the largest city in the UK, a
one year lag of the website density in the nearest city and the same for
the nearest retail centre. Due to the small sizes of the retail 
centres, the latter is only relevant for and included in the OA models and not 
in the LAD models. In addition, the model includes the distance to London, the 
nearest city and the nearest retail centre. The underlying logic is that the 
level of website adoption in a spatial unit depends on the level of the 
adoption in places further up in the urban hierarchy the previous year. The
inclusion of the distance variables incorporates spatial structure into
the hierarchy argument. To depict the neighbourhood mechanism, the
website density of the neighbouring spatial units in the previous year
is employed. This is calculated using rook continuity. Again, the
underpinning rationale is that the level of web adoption within a
spatial unit depends on the level of web adoption in the neighbouring
spatial units the year before. This represents the 'hitting nearby
locations first' argument. Therefore, the spatial and temporal lag of
the website density in LAD and OA is calculated. Lastly, the temporal
pattern of the cumulative adoption is captured by a time trend variable.
Hence, the model will follow the following generic form (Eq. 3):

Regarding the growth model point, growth rates can be and have been calculated
and modelled for LAD. However, such a strategy would have been problematic for 
OA, because of the very large frequency of OA without any websites. As 
mentioned in the text:

> In 1997 only 1% of the UK OA were associated
with at least one website. This should not come as a surprise given that this
was the very beginning of the commercial Internet and any activities
with a digital footprint were concentrated in a handful of areas, as 
illustrated in Figure 5. At the end of the study
period almost half of the UK OA were not associated with a website.
Again, given the granularity of the data this was expected.

Therefore, if it was to to calculate growth rates for OA,  a small constant
should have been added to all OA. Given how many 0s exist, this would have been
problematic and add unnecessary noise. Given that a multi-scale and granular
approach was important for this paper and part of its novelty, the levels 
modelling framework was adopted. However, the current revised version of the 
paper now reports such a LAD growth model in an appendix. The following text has 
been added in the methods section:

> An alternative specification could model the growth of website density. However, 
such a strategy is problematic because of the large frequency of 
OA without websites, which is to be expected given their small size and high 
granularity. Still, as a robustness text, the results of a growth model for
LAD are presented in the Appendix.

Then, the following text has been added in the Results section:

> In addition, Table A3 in the Appendix
presents the accuracy results of a LAD model the dependent variable of which 
is the yearly growth rate of website density. Even for such a specification
(see Equation 7 in the Appendix) the predictive capacity of
model is exceptionally high ($R^2 = 0.5$).

@@

\color{teal}
\noindent 5. Results\
\noindent 1. Diffusion Speed: The decision to binarize diffusion speed is questionable. Continuous measures, such as growth rates or the time required to reach 50% saturation, would provide more informative insights.
\color{black}

Please see my response to the previous point regarding modelling growth. 
Specifically for this point, the map with the 'binarised' speed (now Figure 8) 
has been moved further down (Section 4.2) to address H~2~. Please note that 
the modelling part does not use this binarised variable, but instead continuous 
ones. Regarding the map, its aim is to depict heterogeneity in Web adoption. I 
could have, of course, plotted the year that each LAD reached their inflection 
point, which is very similar to "the time required to reach 50% saturation", 
using 4-5 bins. But I think the comparison with the country average makes it 
more accesible for the reader. The following text has been added in the first 
paragraph of section 4.2:

> Inflection points have been used in the relevant literature to compare diffusion processes across different settings and technologies [@cherp2021national; @woo2017exploring] as they can enable more insightful comparisons than, for example, the average growth over the 1996-2012 study period.

In addition, the inflection ($t_0$) estimates, which used for the above map 
are presented in detail in the appendix and are also discussed in the main body 
of the paper:

> Importantly, there is obvious spatial clustering of fast and slow LAD. Out of the ten fastest LAD based on when they reached their inflection point (see detailed Table A.2 in the Appendix), nine were located in the South East of England and London and one in Cambridge.

They are also presented in Table 6 for the late adopters discussion.

@@

\color{teal}
\noindent 2. Geographical Scale: While the effort to model diffusion curves at a fine geographical scale is commendable, the paper does not sufficiently explain the benefits or additional knowledge gained from this approach.
\color{black}

A new figure (Figure 3) and the following text have been added to address this
point.

> What escaped from previous studies was εexploring whether such a pattern also 
describes cumulative adoption of new technologies at sub-national scales. To 
test this, the same  logistic function is estimated for each of the 374 LAD. The
detailed results are presented in Table \ref{table.s.lads} in the Appendix. 
The key finding is that for 342 LAD (91% of all LAD) the $R^2 > 0.9$ for the 
logistic function, which indicates that even at local scale the cumulative 
adoption of the Web is characterised by a pattern similar to the one observed 
for countries. Graphically, all these LAD curves are plotted in Figure 
\ref{s_lad_uk}. The different colour lines illustrate whether a LAD reached $t_0$ 
faster or slower than the UK in total. This heterogeneity is further discussed 
later in this section.

> This is the first time that such cumulative adoption curves have been estimated
for areal units as small as the UK LAD. Understanding and modelling the dynamics
of technological diffusion and, ultimately, being able to predict diffusion 
rates at such small local scales can directly support the deployment of 
infrastructural networks (e.g. 5G) and local technology policies. The above are 
in line with previous studies exploring the diffusion of web technologies 
[@PAPAGIANNIDIS2015308], social media [@lengyel2020role] and online
shopping [@bakher2013diffusion].

@@

\color{teal}
\noindent 3. Early Adopters and Latecomers: The discussion of early-adopter and latecomer advantages is underdeveloped. How do these advantages manifest locally? What specific benefits arise if a company registers its website six months earlier than competitors in other locations? If this aspect is retained, it should be elaborated in the theory section, including how it relates to local capabilities or diffusion hierarchies.
\color{black}

The results section has been restructured in order to match the two hypotheses
structure. The early adopters part has been further developed with more 
empirical results and now answers the second hypothesis (H~2~). The key message
is that the diffusion of the Web at the detailed scale of LAD is 
characterised by spatial heterogeneity, well-established spatial patterns and 
both stability and volatility. Although the relevant literature neglected local
scales, the above findings illustrated that such local dynamics exists and 
worth further consideration and investigation especially regarding potential 
local economic effects.

@@

\color{teal}
\noindent 4. Descriptive Statistics: The value of the descriptive Moran’s I and Gini coefficient analyses in Section 4.2 is unclear. Instead, calculating growth rates of adoption and examining their distribution over time would offer a better sense of adoption speed and magnitude.
\color{black}

Please see previous responses about growth rates, zeros and OA. Specifically for 
this point, Moran's I and the Gini coefficient offer some first insights about the
two of the three diffusion mechanisms that H~1~ tests. Section 3 has now been 
restructured based on the two hypotheses, and the following text describes their
value:

> To assess the second diffusion mechanisms, Figures 4 and 5  offer a first insight into whether a neighbourhood or, in other words, a distance mechanism underpins the diffusion of the Web in the UK as they present the Moran's I and the LISA maps of website density respectively for LAD and OA.

More broadly, on top of the fact that in order to calculate growth rates for OA 
a small constant should have been added to overcome the 0s problem, it is not 
clear how their distribution could offer insights about the two diffusion 
mechanisms. It is also unclear to me how such distribution could be used -- the 
distribution of yearly growth rates for every LAD and OA or for all LAD and OA? 
Importantly there are c. 230,000 OA.

@@

\color{teal}
\noindent 5. Retail Centers: The sudden shift to retail centers on page 15 is confusing, given that the data pertains to all commercial sites. Clarify the rationale for this focus.
\color{black}

The retail centres discussion has now been moved in the methods sections, so it
appears much earlier in the paper. In addition, the following text has been 
modified in the relevant paragraph:

### 2025-03-12 paper/submission2/response.qmd

> To capture the hierarchical mechanism the model include as predictors a
one year lag of website density in London, the largest city in the UK, and a
one year lag of the website density in the nearest city. Both variables clearly
illustrate the urban hierarchy in the UK. Especially for the OA models, a 
third predictor is added to capture hierarchy at a much more granular scale:
a one year lag of the website density in the the nearest retail centre [^4]. As
the literature illustrates, retail centres are key components of urban systems
[@dennis2002central, @jones2021reframing]. Due to the small sizes of the retail 
centres, the latter is only relevant for and included in the OA models and not 
in the LAD models.

> [^4]: The retail centres data have been obtained by the Consumer Data Research Centre, an ESRC data investment, under project ID CDRC 498-01, ES/L011840/1; ES/L011891/1: <https://data.cdrc.ac.uk/dataset/retail-centre-boundaries-and-open-indicators>.

@@

\color{teal}
\noindent 6. Modeling Details: The methodological details in Section 4.3 should be moved to the methods section to maintain clarity and coherence. Including these details in the results section is confusing.
\color{black}

The modelling details have now been moved to the methods section.

@@

\color{teal}
\noindent 7. Clarity and Structure: Section 4.3 is difficult to follow. Clearly state the main message and highlight how diffusion mechanisms vary spatially. Develop hypotheses to structure the results and provide potential explanations for observed heterogeneity. Without this, the section feels like a descriptive summary without clear takeaways for the reader.
\color{black}

This section has been restructured based on the two hypotheses. I believe the 
key messages are now more clearly communicated.

@@

\color{teal}
\noindent 
Overall Recommendations
\noindent 1. Consistency: Ensure consistent terminology and focus throughout the paper, particularly when referring to the web and commercial websites.
\noindent 2. Structure: Adopt a clearer structure by separating the introduction, theory/literature review, methods, and results sections.
\noindent 3. Clarity: Develop hypotheses to frame the results and provide clearer explanations for methodological choices and findings.
\color{black}

Thank you, this was very helpful. The whole paper has been restructured based 
on the above suggestions.
